# Part II: R Fundamentals {-}

Before starting, here are some Official manuals and books we will be learning from :

1. https://cran.r-project.org/doc/manuals/r-release/R-intro.html
1. https://cran.r-project.org/doc/manuals/r-release/R-data.html
1. https://cran.r-project.org/doc/manuals/r-release/R-exts.html
1. https://cran.r-project.org/doc/manuals/r-release/R-lang.html
1. [R programming for Data Science by Roger D. Peng.](https://bookdown.org/rdpeng/rprogdatascience/)
1. [R for Data Science](https://r4ds.had.co.nz/)
1. [Intro R book](https://katrienantonio.github.io/intro-R-book/)
1. [R workshop](https://unl-statistics.github.io/R-workshops/)
1. [Intro to R](https://intro2r.com/)


### R as a calculator {-}

The best way to get used to R is to use it as a calculator.

You can start by using the R console and doing simple operations. 

### Basic arithmetic operations {-}

- Addition

```{r}
3+5
```

- Subtraction

```{r}
143-12
```

- Multiplication

```{r}
4*5
```

- Division

```{r}
180/23
```

- Exponientiation

```{r}
4^2
```

#### Arithmetic Functions {-}

this functions may be useful to replace your calculator

- Absolute value 

```{r}
abs(-23)
```

- Square root

```{r}
sqrt(16)
```

- Remainder/modulo 

```{r}
7 %% 3
```

- Logarithms and exponentials 
i.e, 

$$\log_a b = c,\quad ln_e b=a , \quad  e^{a}=b$$
```{r}
log2(4)
log10(1000)
log(4)
exp(8)
2.71828^8
```


### Variables and data types (numeric,character) {-}

#### Assigment Operators {-}
 in `R`, to create a variable , you can use the assigment symbol `<-`, or `=` however, the later is not commonly used in `R`. 
 
 To assign the value of 7 to x , we will do 
 
```{r}
x <- 7
print(x)
```
 
We can now perform operations on these variables
```{r}
3*x+3 # 3*7+3 = 21+3 =24
```

**Remark:** R is case sensitive so, `x` is different from `X`

calling `print(X)` will output error
```{r, error=TRUE}
print(X)
```

#### Data types {-}

R has five basic or "atomic" classes of objects:

- character
- numeric (real numbers) 
- integer
- complex
- logical (True/False)


#### Exercise 1 {-}

Run the following code, then use typeof(), class() functions to find out the data type and/or class object. 

```{r}
my_numeric <- 42.5
John_jay <- "university"
my_logical <- TRUE
my_date <- as.Date("05/29/2018", "%m/%d/%Y")
```

Use `typeof()` or `class()`function to find out what the data type of a variable is .

#### What is the difference between typeof() and class()? {-}


Here we can see that `typeof(my_date)` is a double, and `class(my_date)` is `Date`. This is because `typeof` output the lowest level data type of the object. While `class` outputs the class of the object. 

If you are writing code that involves checking whether an element is of an specific data type , then you need to be careful on how you check that. Depending on the function , it may give you a true value when in reality you want a false value to be returned. 

For example, Imagine you are asked to check if all dates in a dateframe have the correct data type. 

In some cases it might be `"05/29/2018"` but in a rare case (maybe due to a data entry error), there is a `"42.5"`

```{r}
typeof(my_date) == typeof(my_numeric)
```

In the previous comparison , it returns true, meaning that the two data types are the same, maybe youu thought you are comparing if both are date type, but in reality , you are comparing the lowest level data types which are indeed equal (double)

Instead, you should do.

```{r}
class(my_date) == class(my_numeric)
```


**Character Data type ** 
A character stores character values or strings 
```{r}
char <- "This is a character data type"
char
typeof(char)
```


**Numeric Data type ** 

This is for numerical values .

```{r}
num <- 3
print(num)
num_2 <- -2.35
num_2
typeof(num_2)
class(num_2)
```

**Integer Data type** 

For integers, we must specify it as so, and to do it , we must convert the data type. Remark: if it is a decimal, it will remove all decimal, acting as a floor function . 

```{r}
int <- as.integer(3.6332)
int
typeof(int)
int2 <- as.integer(7)
int2
typeof(int2)
class(int2)
```

We can also create a integer by adding an L after it 

```{r}
int3 <- 8L
int3
```

Remark: It does not work with decimals 

```{r, error=TRUE}
int4 <- 3.4546L
int4
```

**Complex Data type** 
Complex data types are stored as `x+yi` , i.e, with the imaginary component

```{r}
compl <- 13+7i
compl
typeof(compl)
complex(real = 23, imaginary = 7)
```


**Boolean Data type ** 
This stores boolean values of `TRUE` and `FALSE`

```{r}
my_bool <- TRUE
my_bool

typeof(my_bool)

my_boolean <- F
my_boolean
typeof(my_boolean)

```



#### Converting Data types {-}

**Convert into Numeric**

We can convert values as numeric. Using `as.numeric()` to change the type but keeping the values as they are. 
When converting 

- complex: removes the `imaginary` part
- logical: `TRUE` becomes `1` , `FALSE` becomes `0`
- character: to its numerical values, but if it has letters then to `NA`

We can use `is.numeric()` to check if a variable is numeric 

```{r}
# Complex
is.numeric(compl)
number <- as.numeric(compl)
number
is.numeric(number)

#Logical 
is.numeric(my_bool)
number2 <- as.numeric(my_bool)
number2
is.numeric(number2)

# Character
char
is.numeric(char)
number3 <- as.numeric(char)
number3
is.numeric(number3)

my_char <- "2023"
is.numeric(my_char)
number4 <- as.numeric(my_char)
number4
is.numeric(number4)

```

**Convert into integer**

```{r}
inte1<-as.integer("234")
inte1
typeof(inte1)

inte2<-as.integer(23+6i)
inte2
typeof(inte2)

inte3<-as.integer(F)
inte3
typeof(inte3)
```

**Converting into Logical**

Return `FALSE` for `0` , `TRUE` otherwise
```{r}
print(as.logical(0))
typeof(as.logical(0))

print(as.logical(-324))
typeof(as.logical(-324))
```

#### Exercise 2  {-}

Create 1 datatype of each: Character, numeric, integer, complex, Boolean


### Getting help {-}

You can use the Plots and files pane (bottom left pane) to click on Help and then search for whichever function you need help with. 

You can also use the `?` before each function. 

```{r}
?mean
```
This will open the information about the function on the plots and files pane.



