# Data Structures {-}

## Vectors: Creating, indexing, and operations {-}

```{r}
# Creating a vector
v <- c(1, 2, 3, 4, 5)
print(v)

# Indexing a vector
print(v[2])  # Access the second element

# Vector operations
v2 <- v * 2  # Multiply each element by 2
print(v2)

```

**Exercise**:

1. Create a vector of your favorite numbers.
2. Access the third element in your vector.
3. Create a new vector that is the square of each element in the original vector.

## Data frames: Creating and exploring data frames {-}

```{r}
# Creating a data frame
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Salary = c(50000, 60000, 70000)
)
print(df)

# Exploring data frames
print(dim(df))  # Dimensions of the data frame
print(colnames(df))  # Column names
print(summary(df))  # Summary statistics

```

Exercise:

1. Create a data frame with at least three columns and four rows.
1. Print the number of rows and columns of your data frame.
1. Display summary statistics of your data frame.

## Importing and exporting data (CSV files) {-}

```{r}
# Exporting data to CSV
write.csv(df, "my_data.csv", row.names = FALSE)

# Importing data from CSV
df_imported <- read.csv("my_data.csv")
print(df_imported)

```

