
# Data Structures {-}

## Vectors: Creating, indexing, and operations {-}

```{r}
# Creating a vector
v <- c(1, 2, 3, 4, 5)
print(v)

# Indexing a vector
print(v[2])  # Access the second element

# Vector operations
v2 <- v * 2  # Multiply each element by 2
print(v2)

```

You can give names to the columns of the vector 
```{r}
my_vector <- c("Katrien Antonio", "teacher")
names(my_vector) <- c("Name", "Profession")
my_vector
```

**Exercise**:

1. Create a vector of your favorite numbers.
2. Access the third element in your vector.
3. Create a new vector that is the square of each element in the original vector.

## Matrices {-}

Matrices are vectors with a dimension attribute. The dimension attribute is itself an integer vector of length 2 (number of rows, number of columns)

Matrices are constructed column-wise, so entries can be thought of starting in the “upper left” corner and running down the columns.

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3) 
m
```

Another example
```{r}
my_matrix <- matrix(1:12, 3, 4, byrow = TRUE)
my_matrix
```

Matrices can be created by column-binding or row-binding with the `cbind()` and `rbind()` functions.

```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
```

## Lists {-}

Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R and you should get to know them well. Lists, in combination with the various “apply” functions discussed later, make for a powerful combination.

Lists can be explicitly created using the list() function, which takes an arbitrary number of arguments.

```{r}
x <- list(1, "a", TRUE, 1 + 4i) 
x
```

## Factors {-}

Factors are used to represent categorical data and can be unordered or ordered. One can think of a factor as an integer vector where each integer has a label. Factors are important in statistical modeling and are treated specially by modelling functions like `lm()` and `glm()`.

Using factors with labels is better than using integers because factors are self-describing. Having a variable that has values “Male” and “Female” is better than a variable that has values 1 and 2.

Factor objects can be created with the factor() function.

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no")) 
x
```

Level are put in alphabetical order, but you can also define the levels. 

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"),levels = c("yes", "no"))
x
```

## Data frames: Creating and exploring data frames {-}

Data frames are used to store tabular data in R.

Data frames are represented as a special type of list where every element of the list has to have the same length. Each element of the list can be thought of as a column and the length of each element of the list is the number of rows.

```{r}
# Creating a data frame
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 35),
  Salary = c(50000, 60000, 70000)
)
print(df)

# Exploring data frames
print(dim(df))  # Dimensions of the data frame
print(colnames(df))  # Column names
print(summary(df))  # Summary statistics

```

Exercise:

1. Create a data frame with at least three columns and four rows.
1. Print the number of rows and columns of your data frame.
1. Display summary statistics of your data frame.

Exercise:

Inspect a built-in data frame
```{r}
mtcars
str(mtcars)
head(mtcars)
```

Extract a variable from a data frame and ask a summary 
```{r}
summary(mtcars$cyl) # use $ to extract variable from a data frame
```

Now inspect a tibble

```{r}
library(ggplot2)
diamonds
str(diamonds)  # built-in in library ggplot2
head(diamonds)
```


Can you list some differences?


## Importing and exporting data (CSV files) {-}

```{r}
# Exporting data to CSV
write.csv(df, "my_data.csv", row.names = FALSE)

# Importing data from CSV
df_imported <- read.csv("my_data.csv")
print(df_imported)

```

