
# Part V: Basic Data Visualization (20 minutes) {-}

## Creating simple plots using plot() {-}

Example: Simple scatter plot

```{r}

x <- 1:10
y <- rnorm(10)
plot(x, y, main = "Simple Scatter Plot", xlab = "X Axis", ylab = "Y Axis", col = "blue")

```


## Example 1 {-}

To create a basic scatter plot, you can use the mtcars dataset, which comes built into R. This dataset contains various characteristics of 32 automobiles.

```{r}

data(mtcars)

# 
plot(mtcars$hp, mtcars$mpg, main="MPG vs. Horsepower",
     xlab="Horsepower", ylab="Miles Per Gallon",
     pch=19, col="blue")

```

## Example 2 {-}

Using the pressure dataset, which is also built into R, you can create a simple line plot. The pressure dataset shows the temperature and resulting vapor pressure of mercury.

```{r}

data(pressure)

# Create a line plot
plot(pressure$temperature, pressure$pressure, type="l",
     main="Vapor Pressure of Mercury",
     xlab="Temperature", ylab="Pressure",
     col="red", lwd=2)

```

## Example 3 {-}

```{r}

data(iris)

# Plot Sepal.Length vs. Sepal.Width colored by Species
plot(iris$Sepal.Length, iris$Sepal.Width, col=iris$Species,
     main="Iris Sepal Measurements",
     xlab="Sepal Length", ylab="Sepal Width",
     pch=19)
legend("topright", legend=levels(iris$Species), col=1:3, pch=19)

```

## Exercise {-}
Now, try creating your own visualization using the iris dataset. Here's what you can do:

1. Create a scatter plot using Petal.Length and Petal.Width from the iris dataset.
2. Color the points based on the Species column to differentiate between the species.
3. Add a title, x-axis label, and y-axis label to your plot.
Include a legend that indicates which color corresponds to which iris species.

```{r, echo=FALSE}

data(iris)

# Create the scatter plot
plot(iris$Petal.Length, iris$Petal.Width, col=as.factor(iris$Species),
     main="Iris Petal Measurements",
     xlab="Petal Length", ylab="Petal Width",
     pch=19)

# Add a legend
legend("topright", legend=levels(iris$Species), col=1:length(levels(iris$Species)), pch=19)

```


```{r}
# Load the mtcars dataset
data(mtcars)

# Plot without converting cyl to factor
plot(mtcars$disp, mtcars$mpg, col=mtcars$cyl,
     main="MPG vs. Displacement Colored by Cylinders (Not Factor)",
     xlab="Displacement", ylab="Miles Per Gallon",
     pch=19)
legend("topright", legend=unique(mtcars$cyl), col=unique(mtcars$cyl), pch=19, title="Cylinders")

```

```{r}
# Plot with cyl converted to factor
plot(mtcars$disp, mtcars$mpg, col=as.factor(mtcars$cyl),
     main="MPG vs. Displacement Colored by Cylinders (Factor)",
     xlab="Displacement", ylab="Miles Per Gallon",
     pch=19)
legend("topright", legend=levels(as.factor(mtcars$cyl)), col=1:length(levels(as.factor(mtcars$cyl))), pch=19, title="Cylinders")
```


For later 

## Creating more complex plots ggplot() {-}




```{r}
# install.packages("ggplot2")
library(ggplot2)

# Example: Complex scatter plot with ggplot2
ggplot(data.frame(x, y), aes(x, y)) + geom_point() + 
  ggtitle("Complex Scatter Plot with ggplot2") +
  xlab("X Axis") + ylab("Y Axis")
```

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy, color = class)) +geom_point()
```

## Creating plots with other library {-}
```{r}

# install.packages("lattice")
library(lattice)

# Example: Lattice plot
xyplot(y ~ x, data = data.frame(x, y), main = "Lattice Plot", type = "p")


```

## Creating demographic, map, 3d plots {-}

Installing and loading plotly for 3D plots

```{r}
#install.packages("plotly")
library(plotly)

# Example: 3D Scatter Plot with plotly
plot_ly(data.frame(x, y, z = rnorm(10)), x = ~x, y = ~y, z = ~z, type = "scatter3d", mode = "markers")

# Installing and loading maps package for map plots
#install.packages("maps")
library(maps)

# Example: Map plot
map("world", fill = TRUE, col = "#D3D3D3", border = "white", xlim = c(-130, -60), ylim = c(20, 50))
```


Creating a 3D surface plot with plotly

```{r}

# install.packages("plotly")
library(plotly)

# Simulating a surface for a function (e.g., Rosenbrock function)
rosenbrock <- function(x, y) { (1 - x)^2 + 100 * (y - x^2)^2 }
x <- seq(-2, 2, length.out = 50)
y <- seq(-1, 3, length.out = 50)
z <- outer(x, y, rosenbrock)

# Example: 3D Surface Plot
plot_ly(x = ~x, y = ~y, z = ~z, type = "surface", 
        contours = list(z = list(show = TRUE, usecolormap = TRUE, highlightcolor = "#ff0000", project = list(z = TRUE))))

```



```{r}
# Assuming you've loaded ggplot2 and have it attached
data(diamonds)

# Without as.factor() - Not needed here since `cut` is already a factor
ggplot(diamonds, aes(x = carat, y = price, color = cut)) + 
  geom_point() + 
  theme_minimal() +
  labs(title = "Diamond Price by Carat and Cut", x = "Carat", y = "Price")

# This plot should show distinct colors for each diamond cut

```
```{r}
# Create a numeric version of cut for demonstration
diamonds$cut_numeric <- as.numeric(diamonds$cut)

# Without converting cut_numeric to a factor
ggplot(diamonds, aes(x = carat, y = price, color = cut_numeric)) + 
  geom_point() + 
  theme_minimal() +
  labs(title = "Diamond Price by Carat (Numeric Cut Representation)", x = "Carat", y = "Price")

# With converting cut_numeric to a factor
ggplot(diamonds, aes(x = carat, y = price, color = as.factor(cut_numeric))) + 
  geom_point() + 
  theme_minimal() +
  labs(title = "Diamond Price by Carat (Categorical Cut Representation)", x = "Carat", y = "Price")



```
