# Part III: Building Predictive Models (30 minutes){-}
## Introduction to machine learning in R {-}

*Brief overview of machine learning:* Machine learning in R involves using statistical techniques to enable computers to improve at tasks with experience. It encompasses a variety of techniques for classification, regression, clustering, and more.

```{r}
# Load necessary libraries
#install.packages("caret")
library(caret)


# Example: Splitting a dataset into training and testing sets
data(iris)
set.seed(123) # Setting seed for reproducibility
trainingIndex <- createDataPartition(iris$Species, p = 0.8, list = FALSE)
trainingData <- iris[trainingIndex, ]
testingData <- iris[-trainingIndex, ]


```

*Exercise*:

1. Load a different dataset and partition it into training and testing sets.

## Creating predictive models with caret {-}

```{r}

# Example: Building a predictive model for the iris dataset
model <- train(Species ~ ., data = trainingData, method = "rpart")
print(model)

# Predicting using the model
predictions <- predict(model, testingData)
confusionMatrix(predictions, testingData$Species)

```

*Exercise*:
2. Build a predictive model for another dataset and evaluate its performance.


```

